---
title: "Take Home Exercise 2"
description: VAST Challenge 2025 Mini-Challenge 1
date: 2025-05-13
date-modified: "last-modified"
execute:
  warning: false
  freeze: true
format: 
  html:
    number-sections: false
---

# Overview

For this take-home exercise, we will be working on **Mini-Challenge 1** from the [**VAST Challenge 2025**](https://vast-challenge.github.io/2025/index.html). The objective of this exercise is to design visual analytics methods to track the emergence of a rising artist and forecast the next potential breakout star.

We will address the following questions from the mini-challenge:

#### 1. The Profile of Sailor Shift's Career :

-   Who has Sailor Shift been most influenced by over time?
-   Who has she collaborated with and directly or indirectly influenced?
-   How has she influenced collaborators within the broader **Oceanus Folk** community?

#### 2. The Influence of Oceantus Folk to the Music World :

-   Was this influence **intermittent** or did it show a **gradual rise**?
-   What **genres** and **top artists** have been most influenced by Oceanus Folk?
-   On the conveerse, how has Oceantus Folk changed the rise of Sailor Shift? From which **genre** does it draw most of its contemporary inspiration?

#### 3. Profile of What it Means to be a Rising Star in the Music Industry

-   Visualize the careers of **3 Artist**
-   Compare and Contrast their rise in **Popularity** and **Influence**
-   Using this characterization, gives **3 Predictions** of the next Oceanus Folk stars with be over the next **5 Years**

# The Data 

We will use the dataset provided in the VAST Challenge 2025 Mini-Challenge 1.

::: {.panel-tabset .nav-pills}
#### File Overview

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'

library(knitr)
library(tibble)

dataset_info <- tibble(
  Attribute = c("File Name", "Graph Type", "Number of Nodes", "Number of Edges", "Connected Components"),
  Value = c("MC1_graph.json", "Directed Multigraph", "17,412", "37,857", "18")
)

kable(dataset_info)
```

#### Node

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'
node <- tibble(
  Node_Type = c("Person", "Song", "RecordLabel", "Album", "MusicalGroup")
)

kable(node)
```

#### Edge

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'
edge <- tibble(
  Edge_Type = c(
    "MemberOf", "PerformerOf", "ComposerOf", "ProducerOf", "LyricistOf",
    "InStyleOf", "InterpolatesFrom", "CoverOf", "LyricalReferenceTo",
    "DirectlySamples", "RecordedBy", "DistributedBy"
  )
)

kable(edge)
```
:::

# Setting Up 

## Loading Packages
::: {.panel-tabset .nav-pills}
#### Packages

```{r}
pacman::p_load(jsonlite,tidyverse)
```

#### Packages Information
| **Library** | **Description** |
|-------------------------------|-----------------------------------------|
| [**pacman**](https://cran.r-project.org/package=pacman) | A package management tool that helps load, install, and manage R packages more easily. |
| [**tidyverse**](https://www.tidyverse.org/) | A collection of R packages designed for data science, sharing an underlying design philosophy, grammar, and data structures. |
| [**jsonlite**](https://cran.r-project.org/package=jsonlite) | A robust and flexible R package for parsing and generating JSON data, commonly used for working with web APIs and structured data files. |

:::

::: {.scroll-output}
## Loading Data 
```{r}
data <- fromJSON("data/MC1_graph.json")
glimpse(data)
```
:::

# Data Cleaning

## Extracting Edges

```{r}
edges <- as_tibble(data$links) %>% 
  distinct() 
glimpse(edges)
```
## Extracting Nodes
```{r}
nodes <- as_tibble(data$nodes) %>%
  distinct()

glimpse(nodes)
```


# References 
[**VAST Challenge Kickstarter 2**] (https://isss608-ay2023-24apr.netlify.app/vast/kickstarter2)

