---
title: "Hands on Exercise 4a"
description : Visual Statistical Analysis
date: 2025-04-29
date-modified: "last-modified"
execute:
  warning: false
  freeze: true
format:
  html:
    number-offset: 5
---

# Visual Statistical Analysis

## Learning Outcome

In this hands-on exercise, we will gain hands-on experience on using:

-   ggstatsplot package to create visual graphics with rich statistical information,

-   performance package to visualise model diagnostics, and

-   parameters package to visualise model parameters


## Visual Statistical Analysis with **ggstatsplot**

[**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/index.html) is an extension of [**ggplot2**](https://ggplot2.tidyverse.org/) package for creating graphics with details from statistical tests included in the information-rich plots themselves.

-   To provide alternative statistical inference methods by default.
-   To follow best practices for statistical reporting. For all statistical tests reported in the plots, the default template abides by the [APA](https://my.ilstu.edu/~jhkahn/apastats.html) gold standard for statistical reporting. For example, here are results from a robust t-test:

## Getting Started

### Installing and launching R packages

In this exercise, **ggstatsplot** and **tidyverse** will be used.

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

### Importing data

```{r}
exam <- read_csv("data/Exam_data.csv")
head(exam)
```

### One-sample test: *gghistostats()* method

In the code chunk below, [*gghistostats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/gghistostats.html) is used to to build an visual of one-sample test on English scores.

```{r}
#| code-fold: true
#| code-summary: "Show Code"

set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```