---
title: "Hands on Exercise 1"
description : A Layered Grammar of Graphics ggplot2 methods
date: 2025-04-18
date-modified: "last-modified"
---

# A Layered Grammar of Graphics: ggplot2 methods

## Install and Launch R Packages

```{r}
pacman::p_load(tidyverse)
```

## Importing Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Plotting A Simple Bar Chart

```{r}
ggplot(data = exam_data,aes(x=RACE)) +
  geom_bar()
```

## R Graphics VS ggplot()

::: panel-tabset
### R Graphics

```{r}
hist(exam_data$MATHS)
```

### ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```
:::

## Grammar of Graphics

-   **Data**: The dataset being plotted.

-   **Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

-   **Geometrics**: The visual elements used for our data, such as point, bar or line.

-   **Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

-   **Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

-   **Coordinate systems** define the plane on which data are mapped on the graphic.

-   **Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

## ggplot2:data

```{r}
ggplot(data=exam_data)
```

::: callout-note
### Note

-   A blank canvas appears.
-   `ggplot()` initializes a ggplot object.
-   The `data` argument defines the dataset to be used for plotting.
-   If the dataset is not already a data.frame, it will be converted to one by `fortify()`.
:::

## ggplot2: Aesthetic mappings

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

::: callout-note
### Note

-   ggplot includes the x-axis and the axis’s label.
:::

## ggplot2: geom

Geometric objects are the actual marks we put on a plot. Examples include:

-   geom_point for drawing individual points (e.g., a scatter plot)
-   geom_line for drawing lines (e.g., for a line charts) - geom_smooth for drawing smoothed lines (e.g., for simple trends or approximations)
-   geom_bar for drawing bars (e.g., for bar charts)
-   geom_histogram for drawing binned values (e.g. a histogram)
-   geom_polygon for drawing arbitrary shapes
-   geom_map for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).

::: panel-tabset
### geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

### geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

### geom_histogram()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()      
```

### geom-density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density() 
```

### geom_boxplot()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()  
```

### geom_boxplot() : notches

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE) 
```

### geom_violin()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()     
```
:::

### combined geom objects

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)           
```

## ggplot2: stat

### Working with Stat()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)           
```

### Working with Geom()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)                 
```

## Best Fit Curve for Scatterplot

geom_smooth() is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)      
```

The default smoothing line method can be overwritten as :

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## ggplot2: Facets

::: panel-tabset
### facet_wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### facet_grid()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```
:::

## ggplot2: Coordinates

::: panel-tabset
### coord_flip()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### coord_cartesian()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```
:::

-   [`coord_cartesian()`](https://ggplot2.tidyverse.org/reference/coord_cartesian.html): the default cartesian coordinate systems, where you specify x and y values (e.g. allows you to zoom in or out).
-   [`coord_flip()`](https://ggplot2.tidyverse.org/reference/coord_flip.html): a cartesian system with the x and y flipped.
-   [`coord_fixed()`](https://ggplot2.tidyverse.org/reference/coord_fixed.html): a cartesian system with a "fixed" aspect ratio (e.g. 1.78 for a "widescreen" plot).
-   [`coord_quickmap()`](https://ggplot2.tidyverse.org/reference/coord_map.html): a coordinate system that approximates a good aspect ratio for maps.

## ggplot2: themes

::: panel-tabset
### theme_gray()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  theme_gray()
```

### theme_classic()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  theme_classic()
```

### theme_light()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  theme_light()
```
:::
